Containers and R---the Rockerverse and beyond

### 2019-12-16 manuscript.Rmd
add missing references

The R community keeps growing. This can be seen in the number of new packages on CRAN which keeps on rising, but also in the increased number of conferences, online courses, meetups, and unconferences, as well as in the uptake in education and industry increases [REFs].
All this cements the role of R as the _lingua franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of software.
Combining both these topics, the _Rocker Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: demonstrations, prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

@@

Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability opportunity is being provided.
But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment [REFERENCE!].   
While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.

@@

**[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
_pqR_ fixes some opinionated issues in the R language and is the basis for experimental features, e.g. automatic diffentiation\footnote{ \href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.

**FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.

@@

As with any other software, each new version of R comes with its share of changes.
Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different results from another version can be an issue when it comes to reproducibility of workflows and stability of applications.
For example, think about R random seed: starting with `R 3.6.0`, running `set.seed(2811); sample(1:1000, 2)` will not give the same result as if it was run inside an older version of R.
That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change.
Controlling the version of software is crucial for reproducible research [REF].

@@

Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible [REF].
GPUs are increasingly popular for compute-intensive machine learning tasks [REF].
Though in this case, containers are not completely portable between hardware environments.
Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
The Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages: https://github.com/rocker-org/ml.

>>>>

The R community keeps growing. This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

@@

Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability opportunity is being provided.
But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.

@@

**[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g. automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.

**FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.

@@

As with any other software, each new version of R comes with its share of changes.
Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different results from another version can be an issue when it comes to reproducibility of workflows and stability of applications.
For example, think about R random seed: starting with `R 3.6.0`, running `set.seed(2811); sample(1:1000, 2)` will not give the same result as if it was run inside an older version of R.
That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change.
Therefore, controlling the version of software is most crucial for reproducible research \citep[e.g.][]{boettiger_introduction_2015}.

@@

Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
GPUs are increasingly popular for compute-intensive machine learning tasks, e.g. deep artificial neural networks \citep{schmidhuber_deep_2015}.
Though in this case, containers are not completely portable between hardware environments.
Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
The Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages: https://github.com/rocker-org/ml.
